---
title: "sccm demo for ATAA"
author: "Jianing Zhang"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require(remotes)) install.packages("remotes")
require(R.filesets)
require(Seurat)
require(Matrix)
require(dplyr)
require(purrr)
require(tibble)
require(stringr)
require(CellChat) 
# remotes::install_github("sqjin/CellChat")
require(iTALK)
# remotes::install_github("Coolgenome/iTALK")
require(sccm)
# remotes::install_github("zhjning/sccm")

projName = "ATAA"
workDir = file.path("~/Documents/workspace/softwares/0_myRPacks/scCellMediator2024/publish_version/sccm_out/",projName)
dataDir = file.path(workDir, "data")
scriptDir = file.path(workDir, "script")
treatedDir = file.path(workDir, "out")

sapply(c(workDir, dataDir, scriptDir, treatedDir), sccm::createDir)
setwd(workDir)
```

## Load dataset from Single Cell Portal
Import raw 10x dataset and meta. Download the test ATAA dataset from SCP1909. You need to register an SCP account online to download this dataset.
Move or softlink the downloaded dataset to $dataDir.

```{r ATAA_data_import}
dsname = "ATAA" # dataset name
meta.raw = read.delim(file.path(dataDir, "AorticAneurysm_MetaData_V1.txt"))[-1,]
mtx = readMM(file.path(dataDir, "AorticAneurysm_Expression_Matrix_raw_counts_V1.mtx.gz"))
cells = read.delim(file.path(dataDir, "AorticAneurysm_Expression_Matrix_barcodes_V1.tsv.gz"), header=F)
features = read.delim(file.path(dataDir, "AorticAneurysm_Expression_Matrix_genes_V1.tsv.gz"), header=F)
colnames(mtx) = cells[,1]
rownames(mtx) = features[,1]
so = CreateSeuratObject(mtx, assay = "RNA")
meta.raw = meta.raw %>% 
  filter(NAME %in% Cells(so)) %>% 
  column_to_rownames("NAME")
```

```{r Sample_pretreatment}
## pretreating dataset and calling seurat to generate the seurat object for following analysis
so = subset(so, cells = rownames(meta.raw))
so = AddMetaData(so, meta.raw[Cells(so),], colnames(meta.raw)) # so@meta.data[,c("nCount_RNA","nFeature_RNA","percent.mt")] = NULL
so$orig.ident = so$biosample_id
so$CellType = so$celltype
so$Condition = so$disease__ontology_label
solist = separate_so(so = so, batch = "orig.ident")
ct.colorlist = get_random_colorlist(reduce(lapply(solist, function(tmp.so) tmp.so$CellType %>% unique), union))
saveRDS(ct.colorlist, file.path(treatedDir, "ct.colorlist.rds"))
## separating single cells according to the original dataset sources and generating seurat objects for single cells per source
for (i in 1:length(solist)){
  sname = names(solist)[i]
  tmp.so = solist[[i]]
  run_seurat(tmp.so, label = sname, features_to_view = c("orig.ident", "CellType"),
             save = T, so_dir = file.path(treatedDir, dsname), treat.cellname = F,
             mtlabel = "MT-", min_avg = 0.01, assay = "RNA", addNamedColorList = ct.colorlist,
             scale_factor = 1e6, min_bin = 50, min_cutoff = c(0.01, Inf), calcClustMarker = F,
             dimnum = 20, dim_sd = 1.25, dim_pval = 0.01, clustRes = 0.8) # dimnum can be set at 'auto'
}
```

## Study the relationship between CD4+ T cells, CD16+ Monocytes and CD14+ Monocytes 
Generate a combined seurat object for triple-cell causal mediation analysis only including the three cell types to study

```{r triple_cell_subset}
roi_label = "_integrated_VSMC2_Fibroblast_Macrophage/"
celltypeList = c("02. VSMC2","06. Fibroblast","14. Macrophage")
run_combined_seurat(solist_dir = file.path(treatedDir, dsname), ## this folder only allows to store separated datasets rather than the integrated ones
                    save_dir = file.path(treatedDir, paste0(dsname, roi_label)), 
                    ifReturn = FALSE, 
                    filterRules = 'CellType %in% c("02. VSMC2","06. Fibroblast","14. Macrophage") & nFeature_RNA > 500 & nCount_RNA > 1000', 
                    acceptRules = 'CellType,>,20', 
                    dimnum = 20, k.anchor = 5, k.filter = 50, k.weight = 20)
```

## Prepare the dataset and calling seurat to generate the seurat object for following analysis

```{r}
so = load_so(file.path(treatedDir, paste0(dsname, roi_label)))
p = UMAPPlot(so, group.by = "CellType", cols = ct.colorlist, raster = T, label = T, pt.size = 2.5, split.by = "Condition")
savePlot(p, file.path(file.path(treatedDir, paste0(dsname, roi_label))), "umap", "celltype", pWd = 6.3, pHt = 3, "pdf")
## running cellchat to estimate the interactions
data_layers = grep("^data",Layers(so),value =T)[1]
data.input <- GetAssayData(so, assay = "RNA", layer = data_layers)
meta <- data.frame(ct = so$CellType, row.names = Cells(so), sample = so$orig.ident)
ctlist = unique(so$CellType)
## generate potential regulatory relationships
ctlist_to_check <- generate_triple_ct_model(ctlist,
                                            cells_to_fix = named_vec(c(3), "02. VSMC2"),
                                            keep_or_remove = "keep")
```

```{r Determine the function to detect}
## search functions about the cell activation and differentiation of monocytes 
get_functions_to_view("vascular_associated_smooth_muscle_contraction", mod = "inspect", overlapKeywords = TRUE,  species = "human", cat = "C5") %>% unique
func_list = sapply(c("GOBP_REGULATION_OF_VASCULAR_ASSOCIATED_SMOOTH_MUSCLE_CONTRACTION",
                     "GOBP_REGULATION_OF_SMOOTH_MUSCLE_CONTRACTION",
                     "GOBP_VASCULAR_ASSOCIATED_SMOOTH_MUSCLE_CONTRACTION"),
                   function(x) get_functions_to_view(x, mod = "catch", overlapKeywords = FALSE, species = "human", cat = "C5"))
funcToUse = func_list %>% unlist %>% unique 
```

## Cell-cell Communication Analysis using CellChat

```{r running cell chat}
require(jzhang.utils)

cellchatDir = file.path(treatedDir, paste0(dsname, roi_label), "cellchat.out")
createDir(cellchatDir)
customedSymbolList = trans_features(match_geneids(rownames(data.input), species = "human", from = "ENSEMBL", to = "SYMBOL"), from = "ENSEMBL", to = "SYMBOL") #change EnsemblID+Symbol to Symbol
run_cellchat(so, feature.name = "CellType", batch.name = "orig.ident", output = cellchatDir, 
             minCNUM.tot = 5, sampleProp = 0.6, minCNUM.new = 5, runs = 3, species = "human", 
             storeResampledSO = F, suggestedRunTimes = F, customedSymbolList = customedSymbolList)

```

```{r running iTalk}

iTALKDir = file.path(treatedDir, paste0(dsname, roi_label), "iTALK.out")
createDir(iTALKDir)

```

